Information is Not Communicated Uniformly: Evidence from Spoken and Written Corpora

### 2019-01-26 cogsci/position-info.Rmd
second cogsci draft ready

Let us assume that the UID perspective is correct. Then we expect that the entropy curve for every language should be close to an affine, linear function with some noise. Entropy should monotonically and uniformly increase at all word positions across utterances on average. We will find a contradiction with this claim through examining the average entropy at each position of a sentence for both speech and text data.

@@

@@

We instead check the effects of individual features on the embeddings of languages in the different treatments. We computed pairwise cosine similarity between each pair of language vectors within a treatment. For a subset of WALS features which had values entered for most of the languages we obtained from Wikipedia, we used a generalized linear model to see whether the cosine similarity between languages mattered in predicting if the languages shared the same value for a WALS feature. For eight features and $45$ languages, we found this to be true.

@@

# Conclusions 
In this paper, we have characterized and applied a model based on Genzel & Charniak (2002) for distinguishing In doing so, we have provided evidence against the Uniform Information Density hypothesis. Our work indicates that information is not distributed uniformly throughout an utterance regardless of languages, but that languages possess a characteristic information distribution determined by their syntactic, morphological and phonological features.

Our work has focused on words within sentences in both speech and text across typologically diverse languages. Other researchers have explored information rate cross-linguistically as a single number characterizing each language, most notably including Aylett & Turk (2004). That work characterizes information rate as information per unit time, and is complementary to the work we accomplish here.

A large volume of research has indicated the effects of surprisal on fixation duration during eye-tracking studies (Demberg & Keller, 2008) (Boston et al., 2008) (Smith & Lev, 2013). Higher surprisal of a word corresponds to higher fixation duration. The *wrap-up effect* is a popular hypothesis within the eye-tracking community. The hypothesis states that, in written text, sentence-final words are processed more slowly on average then sentence-medial or sentence-initial words, due to readers integrating information from the entire sentence in order to form a final, coherent thought expressed by the sentence, among other reasons. The wrap-up effect is drawn from evidence in English, German and Dutch, all of which are languages with a large final increase in their entropy curve from our study. We aim to see in the future if as a consquence of our work in this paper, the wrap-up effect falls out from the final jump in the entropy curves in some languages.

>>>>

The entropy curve is a consequence of the information-theoretic rate of information transfer in a language, where information is defined as a reduction in uncertainty. Thus the more entropy at a word position in a sentence (Yu et al., 2016) or a sentence index in a text (Genzel & Charniak, 2002), the more information is on average contained around that position. Let us assume that the UID perspective is correct. Then we expect that the entropy curve for every language should be close to an affine, linear function with some noise. Entropy should monotonically and uniformly increase at all word positions across utterances on average. We will find a contradiction with this claim through examining the average entropy at each position of a sentence for both speech and text data.

[Put in an illustration of what UID predicts?]

@@

[Put in an illustration of the slope sections and treatments here?]

@@

We instead check the effects of individual features on the embeddings of languages in the different treatments. We computed pairwise cosine similarity between each pair of language vectors within a treatment. For a subset of WALS features which had values entered for most of the languages we obtained from Wikipedia, we used a generalized linear model to see whether the cosine similarity between languages mattered in predicting if the languages shared the same value for a WALS feature. For eight features and $45$ languages, we found this to be true. The table below shows the results for the generalized linear model we computed using WALS features and the absolute treatment.

@@

Cosine distance played some role in the feature determination. The top features were $83$, $81$, $95$ and $97$, which all characterize word order. This makes intuitive sense, and indicates that for this small subset of features and subset of languages that the embeddings in the slope space are related to the WALS features. Therefore typological features play a role in determining the positional entropy values for a language. This means that the entropy curves are in some way structured by the syntactic, morphological and phonological features of a language.

# Conclusions 
In this paper, we have characterized and applied a model based on Genzel & Charniak (2002) for distinguishing entropy at each word position within sentences. In doing so, we have provided evidence against the Uniform Information Density hypothesis. Our work indicates that information is not distributed uniformly throughout an utterance regardless of languages, but that languages possess a characteristic information distribution determined by their syntactic, morphological and phonological features.

Our work has focused on words within sentences in both speech and text across typologically diverse languages. Other researchers have explored information rate cross-linguistically as a single number characterizing each language, most notably including Aylett & Turk (2004). That work characterizes information rate as information per unit time, and is complementary to the work we accomplish here. Languages may possess both an overall information transfer rate, showing the average amount of semantic information transmitted per syllable or word, and an entropy distribution indicative of the language's information distribution.

Research has indicated the effects of surprisal on fixation duration during eye-tracking studies (Demberg & Keller, 2008) (Boston et al., 2008) (Smith & Levy, 2013). Higher surprisal of a word is on average predictive of higher fixation duration. The *wrap-up effect* states that when a person reads written text, he or she will process sentence-final words more slowly on average then sentence-medial or sentence-initial words, due to integrating information from the entire sentence to form a final understanding of the sentence's meaning (Stowe et al., 2018) (Kuperman et al., 2010). The wrap-up effect is drawn from evidence in English, German and Dutch, all of which are languages with a large final increase in their entropy curve from our study. We aim to see in the future if as a consequence of our work in this paper, the wrap-up effect falls out from the final jump in the entropy curves in some languages. We would analyze eye-tracking data from languages with low positional entropy in sentence-final positions. Since the lexical surprisal metric in eye-tracking and our positional entropy metric are both measures of log-probability surprisal, then in languages such as Hindi where the sentence-final positional entropy values are low we expect to find no wrap-up effect.

# Acknowledgements

[Not here yet.]

# References

### 2019-01-26 cogsci/position-info.tex
second cogsci draft ready

Let us assume that the UID perspective is correct. Then we expect that
the entropy curve for every language should be close to an affine,
linear function with some noise. Entropy should monotonically and
uniformly increase at all word positions across utterances on average.
We will find a contradiction with this claim through examining the
average entropy at each position of a sentence for both speech and text
data.

@@

The UID hypothesis predicts that information transmission rate will tend
towards uniformity in both speech and text. We begin by examining speech
to determine if we obtain the entropy curve shape we expect for spoken
communication. We begin with the same source as Levy \& Jaeger (2007)
and Jaeger (2010), the Switchboard corpus of adult telephone
conversations {[}CITATION{]}. We also examine child-adult conversations
in the CHILDES TalkBank (Brown 1973; MacWhinney 2014) corpora database
of spoken adult-child conversations in multiple languages. Children are
not fully developed speakers, so we want to compare the entropy curve we
obtain by computing over the utterances in the CHILDES corpora to the
utterances in the adult-adult Switchboard corpus.

@@

Using Giuseppe Attardi's Wikiextractor tool
\footnote{https://github.com/attardi/wikiextractor}, we extracted the
text corpora for \(165\) languages from Wikipedia by downloading a
stored collection of Wikipedia entries in each langauge and randomly
selecting several thousand articles from each Wikipedia language. Each
language corpus was cleaned and limited to sentences between \(6\) and
\(50\) words. Similar to the data from CHILDES, we divided each corpus
by sentence length, and then computed the unigram entropy measure on
each word position within each sentence length.

To classify the unigram entropy curves of the different distributions,
we computed three slope treatments of each curve. In the \emph{absolute}
treatment, with sentence length denoted as \(n\), we computed the slope
between positions \(1\) and \(2\), positions \(2\) and \(3\), positions
\(3\) and \(n-2\), positions \(n-2\) and \(n-1\) and positions \(n-1\)
and \(n\). For the short utterances appearing the CHILDES speech corpora
we examined, these appeared to be important junctions in the
distributions, with a seeming plateau in the middle of the unigram
entropy curve for each of the language corpora we examined in CHILDES.

@@

For the German and Japanese entropy curves, we observed the same shape
for the text Wikipedia corpora entropy curves and the speech CHILDES
corpora entropy curves, supporting the argument for robustness across
text and speech.

@@

In the \emph{absolute} and \emph{relative 5} treatments, each language
is cast as a point in \(5\)-dimensional space. In the \emph{relative 10}
treatment, each language is cast as a point in \(10\)-dimensional space.
This allows for direct comparison using cosine similarity and moreover
for unsupervised clustering analysis to compare the results of the
Wikipedia entropy analysis to known typological features within the
langauges in the Wikipedia dataset. Using the R hclust package, we
performed a hierarchical clustering of the Wikipedia slope data within
each treatment. A subset of the results at a glance for the
\emph{absolute} treatment are below.

To determine which phonological, morphological and syntactic features
affected the embedding of a language in the Wikipedia dataset, we looked
at the list of \(144\) linguistic features in World Atlas of Language
Structures (Dryer \& Haspelmath, 2013). We limited the languages in the
WALS database to only those in our Wikipedia dataset and performed
missing-value imputation to obtain the features not coded in WALS for
the languages from Wikipedia.

@@

We instead check the effects of individual features on the embeddings of
languages in the different treatments. We computed pairwise cosine
similarity between each pair of language vectors within a treatment. For
a subset of WALS features which had values entered for most of the
languages we obtained from Wikipedia, we used a generalized linear model
to see whether the cosine similarity between languages mattered in
predicting if the languages shared the same value for a WALS feature. We
found this to be true.

@@

@@

In this paper we have explored the differences in rate of information
transmission among spoken child-adult conversational corpora in several
languages and text corpora pulled from Wikipedia for well over a hundred
other corpora. We have seen that the entropy curve, a representation of
information at each position of an utterance without regard to context,
varies from language to language and is predicted by a subset of
syntactic, phonological and morphological features in a language. We
have provided evidence against the Uniform Information Density
hypothesis and provided an alternate explanation for the rate at which
information is encoded and transmitted in human communication, namely
one that is robust across corpora, language dependent and arises from
the features of the language.

A large volume of research has indicated the effects of surprisal on
fixation duration during eye-tracking studies: higher surprisal of a
word corresponds to higher fixation duration. The \emph{wrap-up effect}
is a popular hypothesis within the eye-tracking community. The
hypothesis states that, in written text, sentence-final words are
processed more slowly on average then sentence-medial or
sentence-initial words, due to readers integrating information from the
entire sentence in order to form a final, coherent thought expressed by
the sentence, among other reasons. The wrap-up effect is drawn from
evidence in English, German and Dutch, all of which are languages with a
large final increase in their entropy curve from our study. We aim to
see in the future if as a consquence of our work in this paper, the
wrap-up effect falls out from the final jump in the entropy curves in
some languages.

>>>>

The entropy curve is a consequence of the information-theoretic rate of
information transfer in a language, where information is defined as a
reduction in uncertainty. Thus the more entropy at a word position in a
sentence (Yu et al., 2016) or a sentence index in a text (Genzel \&
Charniak, 2002), the more information is on average contained around
that position. Let us assume that the UID perspective is correct. Then
we expect that the entropy curve for every language should be close to
an affine, linear function with some noise. Entropy should monotonically
and uniformly increase at all word positions across utterances on
average. We will find a contradiction with this claim through examining
the average entropy at each position of a sentence for both speech and
text data.

{[}Put in an illustration of what UID predicts?{]}

@@

The UID hypothesis predicts that information transmission rate will tend
towards uniformity in both speech and text. We begin by examining speech
to determine if we obtain the entropy curve shape we expect for spoken
communication. We begin with the same source as Levy \& Jaeger (2007)
and Jaeger (2010), the Switchboard corpus of adult telephone
conversations. We also examine child-adult conversations in the CHILDES
TalkBank (Brown 1973; MacWhinney 2014) corpora database of spoken
adult-child conversations in multiple languages. Children are not fully
developed speakers, so we want to compare the entropy curve we obtain by
computing over the utterances in the CHILDES corpora to the utterances
in the adult-adult Switchboard corpus.

@@

Using Giuseppe Attardi's Wikiextractor tool
\footnote{https://github.com/attardi/wikiextractor}, we extracted the
text corpora for \(165\) languages from Wikipedia by downloading a
stored collection of Wikipedia entries in each langauge and randomly
selecting several thousand articles from each Wikipedia language. Each
language corpus was cleaned and limited to sentences between \(6\) and
\(50\) words. Similar to our process for the Switchboard and CHILDES
spoken corpora, we divided each corpus by sentence length, and then
computed the unigram entropy measure on each word position within each
sentence length. We wanted to directly compare and classify the unigram
entropy curves of the languages from Wikipedia. We computed three slope
treatments of each curve. In the \emph{absolute} treatment, with
sentence length denoted as \(n\), we computed the slope between
positions \(1\) and \(2\), positions \(2\) and \(3\), positions \(3\)
and \(n-2\), positions \(n-2\) and \(n-1\) and positions \(n-1\) and
\(n\). For the short utterances appearing the CHILDES speech corpora we
examined, these appeared to be important junctions in the distributions,
with a seeming plateau in the middle of the unigram entropy curve for
each of the language corpora we examined in CHILDES.

@@

{[}Put in an illustration of the slope sections and treatments here?{]}

@@

In the \emph{absolute} and \emph{relative 5} treatments, each language
is embedded in \(5\)-dimensional space. In the \emph{relative 10}
treatment, each language is embedded in \(10\)-dimensional space. This
allows for direct comparison between languages on Wikipedia using cosine
similarity and moreover for unsupervised clustering analysis to compare
the results of the Wikipedia entropy analysis to known typological
features within the langauges in the Wikipedia dataset. To determine
which phonological, morphological and syntactic features affected the
embedding of a language in the Wikipedia dataset, we looked at the list
of \(144\) linguistic features in World Atlas of Language Structures
(Dryer \& Haspelmath, 2013). We limited the languages in the WALS
database to only those in our Wikipedia dataset and performed
missing-value imputation to obtain the features not coded in WALS for
the languages from Wikipedia.

@@

We instead check the effects of individual features on the embeddings of
languages in the different treatments. We computed pairwise cosine
similarity between each pair of language vectors within a treatment. For
a subset of WALS features which had values entered for most of the
languages we obtained from Wikipedia, we used a generalized linear model
to see whether the cosine similarity between languages mattered in
predicting if the languages shared the same value for a WALS feature.
For eight features and \(45\) languages, we found this to be true. The
table below shows the results for the generalized linear model we
computed using WALS features and the absolute treatment.

@@

Cosine distance played some role in the feature determination. The top
features were \(83\), \(81\), \(95\) and \(97\), which all characterize
word order. This makes intuitive sense, and indicates that for this
small subset of features and subset of languages that the embeddings in
the slope space are related to the WALS features. Therefore typological
features play a role in determining the positional entropy values for a
language. This means that the entropy curves are in some way structured
by the syntactic, morphological and phonological features of a language.

@@

In this paper, we have characterized and applied a model based on Genzel
\& Charniak (2002) for distinguishing entropy at each word position
within sentences. In doing so, we have provided evidence against the
Uniform Information Density hypothesis. Our work indicates that
information is not distributed uniformly throughout an utterance
regardless of languages, but that languages possess a characteristic
information distribution determined by their syntactic, morphological
and phonological features.

Our work has focused on words within sentences in both speech and text
across typologically diverse languages. Other researchers have explored
information rate cross-linguistically as a single number characterizing
each language, most notably including Aylett \& Turk (2004). That work
characterizes information rate as information per unit time, and is
complementary to the work we accomplish here. Languages may possess both
an overall information transfer rate, showing the average amount of
semantic information transmitted per syllable or word, and an entropy
distribution indicative of the language's information distribution.

Research has indicated the effects of surprisal on fixation duration
during eye-tracking studies (Demberg \& Keller, 2008) (Boston et al.,
2008) (Smith \& Levy, 2013). Higher surprisal of a word is on average
predictive of higher fixation duration. The \emph{wrap-up effect} states
that when a person reads written text, he or she will process
sentence-final words more slowly on average then sentence-medial or
sentence-initial words, due to integrating information from the entire
sentence to form a final understanding of the sentence's meaning (Stowe
et al., 2018) (Kuperman et al., 2010). The wrap-up effect is drawn from
evidence in English, German and Dutch, all of which are languages with a
large final increase in their entropy curve from our study. We aim to
see in the future if as a consequence of our work in this paper, the
wrap-up effect falls out from the final jump in the entropy curves in
some languages. We would analyze eye-tracking data from languages with
low positional entropy in sentence-final positions. Since the lexical
surprisal metric in eye-tracking and our positional entropy metric are
both measures of log-probability surprisal, then in languages such as
Hindi where the sentence-final positional entropy values are low we
expect to find no wrap-up effect.

\section{Acknowledgements}\label{acknowledgements}

{[}Not here yet.{]}

\section{References}\label{references}
